---
title: "Interactive session 4a"
format: html
editor: visual
---

#### Gotta Love Linear Algebra

```{r}
pacman::p_load("tidyverse")
```

lets make a matrix of 10 values

```{r}
my_vals = seq (from = 1, 
               to = 10, 
               by = 1);my_vals
#and look at it 

```

but I want 2 rows, and R lets me do that

```{r}
my_matrix = matrix(data= my_vals, nrow = 2, ncol = 5 , byrow = T); my_matrix


```

but r is weird and certain things it will not like

```{r}
my_matrix = matrix(data= my_vals, nrow = 2, ncol = 3 , byrow = T); my_matrix

```

it will cut off some

what if theres too much space?

```{r}
my_matrix = matrix(data= my_vals, nrow = 3, ncol = 4 , byrow = T); my_matrix

```

always look what you make

#### Scalar multiplication, addition, subtraction

```{r}
4*my_matrix

```

addition and subtraction need matrices of the same length

```{r}
my_vals2 = seq(from = 21,
               to = 30,
               by = 1)
my_matrix2= matrix(data = my_vals2, nrow = 2, ncol = 5);my_matrix2
```

cool now we can add and subtract

```{r}
#Add
my_matrix+my_matrix2

#similar to subtraction 
my_matrix2 - my_matrix
```

## Matrix Multiplication

To show this we get to make 2 2x2 matrices

```{r}
cats = matrix(data= c(0,4,3,1), nrow = 2, ncol = 2, byrow = T);cats
dogs = matrix(data= c(6, -3, 0, 2), nrow = 2, ncol = 2, byrow = T);dogs
```

just like for vectors you use %\*%

```{r}
cats %*% dogs
```

Nice

#### Leslie Matrix Example 

first we create matrix that we saw in class

```{r}
insect_leslie = matrix (c(0, 0 , 600, .2, 0, 0, 0, .08, 0 ), nrow =3,
                        ncol = 3, byrow = T);insect_leslie
```
looks good 

now lets specify some initial conditions : 12000 eggs, 700 larva, 500 adult females

```{r}
insecty0= c(12000, 700, 500); insecty0
```

Now we can use the dot product to find year1 

```{r}
insecty1= insect_leslie %*% insecty0; insecty1
```
nice now that we have that we can keep going 

```{r}
insecty2= insect_leslie%*%insecty1;insecty2
```
Year 3

```{r}
insecty3 = insect_leslie%*%insecty2; insecty3
```

and lastly year 4

```{r}
insecty4 = insect_leslie %*% insecty3; insecty4
```

#### Another leslie matrix example 

You have collected information about a type of tree. There are three life stages: seeds (S), juvenile (J) and adult (A). Juvenile trees don’t produce seeds. Adults produce 5000 seeds each year. 10% of seeds produced become seedlings (juveniles), and 6% of seedlings go on to become adults. 95% of adult trees survive each year.

Create the Leslie matrix for this population
Given an initial population structure of 5000 seeds, 0 juveniles, and 0 adults, project the population structure over 4 years.

Ok so first we need the tree matrix 

```{r}
tree_leslie = matrix(c(0,0, 5000, .1, 0, 0, 0, .06, .95),
                     nrow = 3,
                     ncol = 3,
                     byrow = T);tree_leslie
```

ok now we need initial conditions 

```{r}
treey0 = c(5000, 0, 0 )
```

cool now lets do the 4 years 

```{r}
#y1
treey1 = tree_leslie%*%treey0; treey1

#y2
treey2 = tree_leslie%*% treey1; treey2

#y3
treey3 = tree_leslie %*% treey2; treey3

#y4
treey4 = tree_leslie %*% treey3; treey4
```
whew that took a while
